\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-15 coqimpl.tex
Save.

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref*{fig:plugin-example2} is definitely not expressive enough. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt <> self_B.ff}.

Propositional Partial Recursor can prove injection and discrimination of the constructors.
Propositional Partial Recursor is good enough because it and its computational axiom can prove that the vanilla inductive type can "embed" into the extensible inductive type. 
(i.e. there will be an left inversion of that injection, which is also witnessing the fact that those "constructors" can really act as constructors) 
Thus 1. every future extension can support this partial recursion; 2. every type support this partial recursor with its computational axiom at least "support these constructors" (because of the embedding).

>>>>

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref*{fig:plugin-example2} is not expressive enough. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to prove injectiveness of constructor we only require a recursor and its computational rules. However, as we pointed out earlier, the recursor of an inductive type cannot be exposed into the context for the sake of extensibility. We may have to alter the formulation of the recursor.

The simplest idea is to allow partiality -- i.e. returning \mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}. We call this \textit{propositional partial recursor}. For example, the propositional partial recursor for \mintinline{Coq}{B.b} is 
\begin{minted}{Coq}
b_prec : forall P : b -> Set, P tt -> P ff 
    -> forall x : b, option (P x). 
\end{minted}
with appropriate computational axioms. The partiality doesn't break extensibility (i.e. all the future extension of "b" can support this "b_prec").

What's more, partial recursor is theoretically enough for the "injection" and "discriminate" tactics. The rough idea to achieve this is to use "b_prec" to reflect our special inductive type into vanilla inductive type, and use the orignal "injection" and "discrimination" respectively. For example, we want to prove \mintinline{Coq}{self_B.tt ≠ self_B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : self_B.b → option bool 
                   := b_prec (fun _ => bool) true false.
Definition tt_neq_ff : self_B.tt = self_B.ff → False.
(* Because self_B.tt = self_B.ff → reflect self_B.tt = reflect self_B.ff 
                → true = false    (by computational axiom)
                → False           (by discriminate of bool) *)
\end{minted}

In fact, the above "reflect" function implies that propositional partial recursor is good enough to some extent, at least for non-indexed inductive type, because we can ``embed'' vanilla inductive type into types supporting partial recursor. Because "reflect" is actually a left inverse of an injection "bool → self_B.b". A bit more formally, and restrict our focus to non-indexed inductve types,

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$
  \begin{itemize}
    \item Define $"C"$ as the inductive type using this list, i.e. 
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} that has ...
    \begin{itemize}
      \item following $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor 
      \begin{minted}{Coq}
prec : forall (R : Set), ..,  
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> R) -> R)), .., D -> R
      \end{minted}
      \item all the following $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    \end{itemize}
  \end{itemize}
  We will have an \textbf{embedding} from "C" to "D", more concretely, we can have \mintinline{Coq}{inj : C -> D} and

  \mintinline{Coq}{linv : D -> option C} defined using eliminator and "prec", (both act similar to identity) such that
  \begin{minted}{Coq}
    forall c : C, linv (inj c) = some c
  \end{minted}
\end{theorem}

We emphasize that in \cref{thm:prec-complete}, each function $"d"_i$ can actually be considered as a constructor, because they can be reflected to real constructors of "C" using "linv". Thus, \cref{thm:prec-complete} implies that (1) every future extension of the inductive type can support this partial recursion (trivially); (2) every type support this partial recursor with its computational axiom at least support these constructors because of the embedding. In other words, \textit{a type (at least) supports these constructors if and only if this type supports the corresponding partial recursor (with the computational axiom)}.  Thus we can argue that the partial recursor gives a sound and complete characterization of all the extension of a given (non-indexed) inductive type.
